Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-21 paper.qmd
Update paper.qmd

---
title: "A Framework for Reproducible Data Science using Nix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  To create a reproducible analysis, it is not enough to write clean,
  well-documented and tested code. One must also ensure that all software
  dependencies are clearly declared and, ideally, provide a simple way to
  install them. Several tools exist for this purpose, such as the
  containerization solution [Docker]{.pkg}. A complete solution, however,
  requires not only managing the computational environment but also
  orchestrating the execution of the analysis itself in a robust and automated
  way. This paper presents a comprehensive framework based on the [Nix]{.pkg}
  package manager to solve both problems. We first introduce the [R]{.proglang}
  package [rix]{.pkg}, which simplifies the creation of reproducible
  environments with [Nix]{.pkg}. We then introduce [rixpress]{.pkg}, a second
  [R]{.proglang} package that leverages these environments to define and execute
  complex, multi-language analytical pipelines. Together, these tools provide a
  powerful, unified approach to achieving high standards of computational
  reproducibility.

@@

Finally, once the execution environment is correctly configured, additional
clarity is needed on how to *run* the project itself. Which packages should be
loaded first? Which scripts should be executed, and in what order? Without clear
documentation or a logically organized project structure, these operational
details become yet another barrier to reproducibility.

A range of tools now exist to help researchers approach the gold standard of
full reproducibility, or to consistently deploy the same development environment
across multiple machines. Let us first consider the most basic step in this
process: listing the software used. In [R]{.proglang}, the `sessionInfo()`
function provides a concise summary of the software environment, including the R
version, platform details, and all loaded packages. Its output can be saved to a
file and included as part of a study’s reproducibility record. Below is an
example output from `sessionInfo()`:

@@

locale:
 LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 LC_ADDRESS=C               LC_TELEPHONE=C            
 LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C

time zone: Etc/UTC
tzcode source: system (glibc)

@@

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later
time) can easily see which version of [R]{.proglang} and which packages (with
their versions) were used. However, reproducing the environment still requires
manually installing the correct package versions. This can be difficult,
especially when packages depend on system-level libraries. For example, the
[nloptr]{.pkg} package requires a precompiled [nlopt]{.proglang} binary, or
alternatively, [cmake]{.proglang} when building from source on Linux or macOS.

A more robust approach than simply listing package versions is to use the
[renv]{.pkg} package. [renv]{.pkg} captures the project’s software state and
writes it to a `renv.lock` file, which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring that others can recreate
the same setup with minimal effort. Below is an example of an `renv.lock` file:

```{=tex}
\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": [
        "lattice"
      ]
    }

    ... lines below omitted ...
  }
}
\end{Code}
```

This lockfile lists each package alongside its version and the repository from
which it was downloaded. Creating it requires only a single command:
`renv::init()`. Others can then reproduce the same package library by running
`renv::restore()`, which installs the exact package versions in an isolated,
project-specific library that does not interfere with the user’s global R
library.

However, [renv]{.pkg} does not restore the version of [R]{.proglang} itself.
Installing the correct R version must therefore be done separately. Similarly,
`renv` does not handle system-level dependencies such as [cmake]{.proglang},
which is required by [nloptr]{.pkg} on Linux and macOS. These dependencies must
be managed manually or through other tools. Further details are discussed in the
[renv]{.pkg} documentation: \url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

Before continuing, it should be noted that other packages exist which provide
similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg} by
@simonsohn2023 which makes it rather easy to install packages as they were on
CRAN at a given date. For example, the code snippet below installs the
[purrr]{.pkg} [@purrr] and [ggplot2]{.pkg} [@ggplot2] packages as they were on
April 4th, 2017:

```{=tex}
\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}
```

Packages installed with [groundhog]{.pkg} are placed in a project-specific
library, ensuring they do not interfere with other versions of the same packages
used in different projects. Since [groundhog]{.pkg} does not install
[R]{.proglang} itself, users must either install the required R version manually
or use the `tolerate.R.version` argument, as illustrated in the example above.
Without this argument, [groundhog]{.pkg} will not proceed with package
installation if the R version does not match the expected one.

An alternative to [renv]{.pkg} is [rang]{.pkg}, developed by @chan2023, which
similarly allows users to install packages as they existed on a specific date.
Another approach is to use the Posit Package Manager, which provides dated
snapshots of CRAN. For instance, to ensure that packages are installed as they
were on June 30, 2023, one could include the following line in the `.Rprofile`
file:

@@

The `.Rprofile` file is read by [R]{.proglang} at the start of each new session,
meaning that every call to `install.packages()` will install packages from the
specified snapshot mirror. However, unless users explicitly manage separate
libraries for different projects, the Posit Package Manager will install all
packages for all projects as they existed on that snapshot date.

The next step toward the gold standard of reproducibility is ensuring not only
that the correct package versions are installed, but also that the correct
version of [R]{.proglang} itself is used. While this can be done manually,
dedicated tools make the process more efficient. One such tool is [rig]{.pkg},
developed by the [R]{.proglang} Infrastructure Team [-@rlib2023]. [rig]{.pkg}
simplifies installing and managing multiple R versions, allowing users to match
the exact version needed for a given analysis. Once the correct R version is
installed, one can then use tools such as [renv]{.pkg}, [groundhog]{.pkg}, or
[rang]{.pkg} to recreate the associated package library.

However, this multi-step process remains manual, error-prone, and
time-consuming, highlighting the need for more comprehensive and automated
solutions to achieve full computational reproducibility.

The final tool for achieving the gold standard of reproducibility is to bundle
the exact version of [R]{.proglang} and its packages within a [Docker]{.pkg}
image. [Docker]{.pkg} is a containerization platform that allows packaging a
*data product* together with all its dependencies into a self-contained image. A
statistical analysis can be viewed as such a data product, requiring a specific
set of software dependencies to run. *Dockerizing* an analysis involves building
an image that installs these dependencies, often using the tools mentioned
earlier, and includes the analysis scripts and data (but data can also be
dynamically made available at run-time, to avoid having to copy it inside the
image). The steps to create this image are specified in a simple text file
called a `Dockerfile`, which defines the structure and content of the
[Docker]{.pkg} image.

@@

The strength of [Docker]{.pkg} lies not only in its ability to encapsulate the
correct versions of [R]{.proglang} and its packages but also in its inclusion of
system-level dependencies. Since [Docker]{.pkg} images are nearly complete Linux
systems, they naturally bundle libraries and tools required by certain R
packages (for example, [nloptr]{.pkg}’s dependency on the [nlopt]{.proglang}
library). This ensures that future replicators of a study have access to the
same software environment, including all necessary system components. Moreover,
these images can be easily shared, enabling seamless reproducibility across
machines and collaborators.

@@

Despite its strengths, [Docker]{.pkg} can be cumbersome for interactive use.
Although containers can be modified at runtime, any changes are lost once the
container stops unless they are explicitly saved. Running graphical applications
from containers is also possible, but it is generally difficult to configure and
tends to work reliably only on Linux systems. For this reason, [Docker]{.pkg} is
most often used to run web-based applications (such as integrated development
environments, IDEs, like the web version of RStudio) inside containers, offering
a reproducible yet flexible environment for interactive work.

Some IDEs provide built-in support for executing code directly within a running
container, allowing a more seamless workflow. However, these setups typically
require additional configuration, and many general-purpose IDEs are not well
optimized for statistical or data science workflows (though Positron may soon
change that).

A common practice is to conduct a study interactively using a standard
installation of [R]{.proglang} and its packages, and only after completing the
analysis, create a `Dockerfile` to allow others to reproduce the results easily.
For instance, the [dockerfiler]{.pkg} package [@dockerfiler] can automatically
generate a `Dockerfile` from an existing `renv.lock` file, simplifying this
process. While this approach enables post-hoc reproducibility, it does not solve
the challenge of consistently deploying the same environment across multiple
machines as co-authors are working on the project. Ideally, analyses would be
developed directly within a containerized environment from the start, ensuring
that the computational setup is reproducible throughout the project lifecycle.

Another challenge of using [Docker]{.pkg} is that its images are effectively
minimal Linux systems, so familiarity with Linux is highly recommended for
writing reliable and efficient `Dockerfile`s. To guarantee that a build
consistently produces the same image, the `Dockerfile` should reference specific
image *digests* rather than general *tags*. In practice, however, tagged
versions are more commonly used, which can introduce potential reproducibility
issues. [dockerfiler]{.pkg} is an R package that lowers the
barrier to entry by automating parts of the `Dockerfile` creation process, but
they cannot fully replace the need for basic Linux knowledge.

A final step toward the gold standard of reproducibility is to automate the
analysis execution using a build system such as [Make]{.pkg}, rather than
relying on ad hoc script execution. Build automation tools enable analyses to be
defined as a series of reproducible, well-ordered steps that can be executed
consistently. Within [R]{.proglang}, a modern and powerful alternative is the
[targets]{.pkg} package by @landau2021, which provides declarative workflows for
reproducible data analysis.

@mcdermott2021 provides a clear example of a scientific study that achieved the
gold standard of reproducibility. The author created an accompanying GitHub
repository^[https://github.com/grantmcdermott/skeptic-priors] containing
detailed instructions to install the required software and execute the analysis.
Examining this repository reveals the use of multiple tools to capture and share
the computational environment:

- Packages and their versions were recorded in an `renv.lock` file;
- A `Makefile` was used to automate the full analysis and compile the paper;
- A `Dockerfile` provided a complete computational environment, including the
  correct version of [R]{.proglang}, enabling easy execution of the entire
  workflow.

Reaching this gold standard, however, is resource-intensive. It requires
learning a tool for managing package versions, mastering [Docker]{.pkg} for the
programming language and other software dependencies, and using a build
automation tool. Importantly, these considerations are not unique to
[R]{.proglang}: similar approaches are needed for [Python]{.proglang} or any
other programming language, and complexity further increases if several
programming languages are needed for a single project.

As an alternative, we present the [Nix]{.pkg} package manager, which supports
all major operating systems and emphasizes reproducible software installation
and building. [Nix]{.pkg} can replace [Docker]{.pkg}, [renv]{.pkg}, and even
build automation tools such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
for [R]{.proglang} users, we developed the [rix]{.pkg} and [rixpress]{.pkgs}
packages, which are introduced and discussed in this article.

>>>>

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  Reproducible analysis requires more than clean, well-documented code: 
  researchers must also manage software dependencies, computational environments, and workflow execution. 
  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles multi-language environments, 
  system dependencies, and pipeline orchestration together. I present a two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each 
  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking. 
  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports 
  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.

@@

Finally, once the execution environment is correctly configured, additional
clarity is needed on how to *run* the project itself. Which packages should be
loaded first? Which scripts should be executed, and in what order? Without clear
documentation or automated orchestration, these operational details become yet
another barrier to reproducibility.

This paper focuses specifically on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. We present a comprehensive framework that addresses both
challenges using the [Nix]{.pkg} package manager, making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing our approach, we survey existing tools and
their limitations to contextualize our contribution.

### Existing approaches to environment management {#sec-existing}

A range of tools now exist to help researchers approach the gold standard of
full reproducibility, or to consistently deploy the same development environment
across multiple machines. Let us first consider the most basic step in this
process: listing the software used. In [R]{.proglang}, the `sessionInfo()`
function provides a concise summary of the software environment, including the R
version, platform details, and all loaded packages. Its output can be saved to a
file and included as part of a study's reproducibility record. Below is an
example output from `sessionInfo()`:

@@

locale:
 LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 ...

@@

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later
time) can easily see which version of [R]{.proglang} and which packages (with
their versions) were used. However, reproducing the environment still requires
manually installing the correct package versions—a process that becomes
particularly difficult when packages depend on system-level libraries. For
example, the [sf]{.pkg} package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} system libraries.
Installing and configuring these dependencies varies substantially across
operating systems and can be prohibitively difficult on some platforms.

#### Package-level solutions: renv and alternatives

A more robust approach than simply listing package versions is to use the
[renv]{.pkg} package. [renv]{.pkg} captures the project's software state and
writes it to a `renv.lock` file, which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring that others can recreate
the same setup with minimal effort. Below is an abbreviated example of an
`renv.lock` file:
```{=tex}
\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": ["lattice"]
    }
    ... additional packages omitted ...
  }
}
\end{Code}
```

This lockfile lists each package alongside its version and the repository from
which it was downloaded. Creating it requires only a single command:
`renv::init()`. Others can then reproduce the same package library by running
`renv::restore()`, which installs the exact package versions in an isolated,
project-specific library that does not interfere with the user's global R
library.

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[renv]{.pkg} documentation:
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

Other packages provide similar functionality with different trade-offs.
[groundhog]{.pkg} by @simonsohn2023 makes it easy to install packages as they
existed on CRAN at a given date:
```{=tex}
\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}
```

[groundhog]{.pkg} places packages in project-specific libraries but does not
install [R]{.proglang} itself. Without the `tolerate.R.version` argument, it
will refuse to proceed if the R version does not match expectations.
[rang]{.pkg}, developed by @chan2023, offers similar date-based package
installation. The Posit Package Manager provides another approach through dated
CRAN snapshots, configured via the `.Rprofile` file:

@@

This approach installs all packages from the specified snapshot date, but unless
users explicitly manage separate libraries for different projects, all projects
will share the same package versions.

These tools represent significant progress in reproducibility, but they share a
common limitation: **none addresses system-level dependencies or supports
multi-language environments**. A project using [R]{.proglang},
[Python]{.proglang}, and system tools like [Quarto]{.pkg} or \LaTeX requires
coordinating multiple package managers ([renv]{.pkg} for R, virtual environments
for Python, manual installation for system tools), each with its own
configuration and potential for conflicts.

#### System-level solutions: Docker

The most comprehensive approach to reproducibility is containerization with
[Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
dependencies—including the operating system, system libraries, programming
languages, and packages—into a self-contained image. A statistical analysis can
be viewed as such a data product. The steps to create an image are specified in
a `Dockerfile`, which defines the structure and content of the [Docker]{.pkg}
image.

@@

The strength of [Docker]{.pkg} lies in its ability to bundle not only the
correct versions of [R]{.proglang} and its packages but also system-level
dependencies. Since [Docker]{.pkg} images are effectively minimal Linux systems,
they naturally include libraries required by packages like [sf]{.pkg}. This
ensures that future replicators have access to the same software environment,
including all necessary system components. Moreover, these images can be easily
shared via registries like Docker Hub, enabling seamless reproducibility across
machines and collaborators.

@@

Despite its strengths, [Docker]{.pkg} presents several challenges for
statistical computing workflows:

1. Poor interactive support: While containers can be modified at runtime,
changes are lost when the container stops unless explicitly saved. Running
graphical applications like RStudio Desktop from containers requires complex X11
configuration that works reliably only on Linux. Instead, web-based IDEs (like RStudio
Server) are used instead, but come with the inherent limitations of web-based solutions.

2. Steep learning curve: [Docker]{.pkg} images are minimal Linux systems, so
writing reliable `Dockerfile`s requires familiarity with Linux system
administration. To ensure reproducibility, `Dockerfile`s should reference
specific image *digests* rather than *tags*, but this practice is rarely
followed. Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate
`Dockerfile` creation but cannot eliminate the need for basic Linux knowledge.

3. Post-hoc reproducibility: A common workflow is to develop analyses
interactively using standard R installations, then create a `Dockerfile`
afterward to enable reproduction. While this achieves post-hoc reproducibility,
it fails to address the challenge of deploying consistent environments across
multiple machines *during* development, a critical need for collaborative
projects.

4. Single-language focus: While [Docker]{.pkg} can support multi-language
environments, orchestrating them requires manual coordination. There is no
standard way to specify that one step should run in a Python environment while
another runs in R.

### The workflow orchestration gap

Even with a perfectly reproducible environment, researchers face another
challenge: reliably executing the analysis workflow. Which scripts run first?
What are the dependencies between steps? Manual execution is error-prone and
poorly documented. Build automation tools like [Make]{.pkg} address this by
defining analyses as ordered, reproducible steps.

Within [R]{.proglang}, the [targets]{.pkg} package by @landau2021 provides a
modern, declarative approach to workflow management. [targets]{.pkg} tracks
dependencies between code and data, caches intermediate results, and only
recomputes steps affected by changes. This dramatically improves efficiency for
complex analyses.

However, [targets]{.pkg} operates within a single R session. While it can call
Python via [reticulate]{.pkg} or execute system commands, all steps share the
same computational environment. For truly polyglot pipelines (where different
steps may require incompatible dependencies or different language versions) this
limitation becomes problematic. One might run [targets]{.pkg} inside a
[Docker]{.pkg} container to ensure reproducibility, but this requires the entire
pipeline to use a single environment.

@mcdermott2021 provides an excellent example of achieving the gold standard of
reproducibility. The accompanying GitHub
repository^[https://github.com/grantmcdermott/skeptic-priors] demonstrates the
state of the art:

- Package versions recorded in an `renv.lock` file
- A `Makefile` automating the full analysis
- A `Dockerfile` providing a complete computational environment

However, achieving this required mastering multiple complex tools: [renv]{.pkg}
for package management, [Docker]{.pkg} for environment containerization, and
[Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
and failure modes. The complexity compounds for multi-language projects—a Python
project would require learning virtual environments or conda, and coordinating R
and Python in a single analysis requires additional tooling.

### A unified approach with Nix

What researchers need is a tool that:

1. Manages complete environments including programming languages, packages,
and system dependencies;
2. Supports multiple languages natively, not through workarounds;
3. Works interactively without the complexity of containers;
4. Provides step-level isolation so different pipeline steps can use
different environments;
5. Ensures bit-for-bit reproducibility through deterministic builds;
6. Remains simple enough for researchers without systems administration
expertise.

The [Nix]{.pkg} package manager provides these capabilities. [Nix]{.pkg} ensures
reproducible software installation by deploying *component closures*—packages
bundled with all their dependencies and transitive dependencies [@dolstra2004nix].
This creates self-contained software environments that work identically across
machines and over time. [Nix]{.pkg} can replace [Docker]{.pkg} for environment
isolation, [renv]{.pkg} for package management, and even [Make]{.pkg} for
workflow orchestration—all within a single, unified framework.

However, [Nix]{.pkg} has a steep learning curve. It is a complex system with its
own purely functional programming language designed to declaratively define how
software is built and configured. While this functional approach ensures
reproducibility, it creates a significant barrier to adoption.

To make [Nix]{.pkg} accessible to researchers, I developed two [R]{.proglang}
packages:

- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function
calls, eliminating the need to learn the [Nix]{.proglang} language. It handles
environment definition including [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, system tools, and their
dependencies.

- **[rixpress]{.pkg}** orchestrates polyglot analytical pipelines using
[Nix]{.pkg} as the build engine. Each pipeline step runs in its own
hermetically sealed environment with automatic caching and dependency tracking.
This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and [Julia]{.proglang} steps coexist
with different dependencies.

Together, these packages provide what existing tools cannot: a single framework
for managing reproducible, multi-language computational environments and
executing complex analytical workflows with step-level isolation. While
[Nix]{.pkg} remains complex for advanced use cases, [rix]{.pkg} and
[rixpress]{.pkg} abstract this complexity for common research workflows, making
deep reproducibility accessible without requiring systems administration
expertise.

The remainder of this paper proceeds as follows. Section 2 introduces the
[Nix]{.pkg} package manager and explains why its functional approach enables
reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
definition for various use cases. Section 4 discusses our `rstats-on-nix` fork
of the official package repository and how it addresses practical limitations.
Section 5 introduces [rixpress]{.pkg} and demonstrates polyglot pipeline
orchestration through a complete example. Section 6 concludes with discussion of
appropriate use cases and future directions.
